Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-06-01 Documents/M1/realizableFermionicM1/sections/imex.tex
Modified c_sch expression.

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$.
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type~A} (\ee{Insert reference to the origin of Type~A schemes}) and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\[\left( 
   \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
   \end{matrix}
\right)\]
where $\hat{A}$ is invertible.
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
%For $i=1,\ldots,s$, the coefficients
%\begin{align*}
%  c_{i0}
%  =&\left\{
%  \begin{array}{cl}
%    1 - \sum_{j=1}^{i-1}c_{ij}  & \qquad \text{for type $A$,} \\
%    1 - \sum_{j=2}^{i-1}c_{ij}  & \qquad \text{for type $ARS$;}
%  \end{array}
%  \right.\\
%  \tilde{c}_{i0}
%  =&\left\{
%  \begin{array}{cl}
%    0  & \text{for type $A$,} \\
%   \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1}  & \text{for type $ARS$;}
%  \end{array}
%  \right.\\
%  c_{i1}
%    =&\left\{
%    \begin{array}{cl}
%      \sum_{l =1}^{i-1} a_{il}b_{l1} & \qquad ~\text{for type $A$,} \\
%      \qquad ~ 0  & \qquad ~\text{for type $ARS$;}
%    \end{array}
%    \right.\\
%    \tilde{c}_{i1}
%    =&\left\{
%      \begin{array}{cl}
%       \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1} & \text{for type $A$,} \\
%       0  & \text{for type $ARS$;}
%      \end{array}
%      \right.
%\end{align*}
%and the rest coefficients were defined as in \cite{hu_etal_2018}.
For type A schemes, the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, the positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=2,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

Importantly, the coefficients $c_{ij}$ in Eq.~\eqref{eq:imexStagesRewrite} satisfy $\sum_{j=0}^{i-1}c_{ij}=1$, which implies the so-called Shu-Osher form.  
Thus, if the expression inside the square brackets in Eq.~\eqref{eq:imexStagesRewrite} --- which is in the form of a forward Euler update with time step $\hat{c}_{ij}\,\dt$ --- is positive for all $i=1,\ldots,s$, $j=0,\ldots,i-1$, positivity of the entire sum on the right-hand side of Eq.~\eqref{eq:imexStagesRewrite} follows from convexity arguments.  
Thus, if the explicit update with the transport operator is positive for a time step $\dt_{\mbox{\tiny Ex}}$, the IMEX scheme is positivity-preserving for a time step $\dt\le c_{\mbox{\tiny Sch}}\,\dt_{\mbox{\tiny Ex}}$, where
\begin{equation}
  c_{\mbox{\tiny Sch}}=\min_{ij}\,\f{1}{\hat{c}_{ij}}.  
  \label{eq:imexCFL}
\end{equation}
It is desirable to make $c_{\mbox{\tiny Sch}}$ as large (close to $1$) as possible.  
(Positivity of $\vect{u}^{(i)}$ also requires the implicit solve to preserve positivity.)  
In \cite{hu_etal_2018}, Hu et al. provide examples of GSA, positivity-preserving IMEX schemes of type A (see scheme PA2 in \ref{app:butcherTables}) and type ARS.  
In \ref{app:butcherTables}, we provide another example of a GSA, positivity-preserving IMEX scheme of type A (scheme PA2+), with a significantly larger $c_{\mbox{\tiny Sch}}$ (a factor of about $1.7$ larger).

>>>>

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$.
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type~A} \cite{pareschiRusso_2005} and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\[\left( 
   \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
   \end{matrix}
\right)\]
where $\hat{A}$ is invertible.
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
%The coefficients are defined as $c_{i0} = 1 - \sum_{j=1}^{i-1}c_{ij} $, 
%For $i=1,\ldots,s$, the coefficients
%\begin{align*}
%  \tilde{c}_{i0}
%  =&\left\{
%  \begin{array}{cl}
%    0  & \text{for type $A$,} \\
%   \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1}  & \text{for type $ARS$;}
%  \end{array}
%  \right.\\
%    \tilde{c}_{i1}
%    =&\left\{
%      \begin{array}{cl}
%       \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1} & \text{for type $A$,} \\
%       0  & \text{for type $ARS$;}
%      \end{array}
%      \right.
%\end{align*}
%and the rest coefficients were defined as in \cite{hu_etal_2018}.
For type A schemes, the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, the positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=2,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

Importantly, the coefficients $c_{ij}$ in Eq.~\eqref{eq:imexStagesRewrite} satisfy $\sum_{j=0}^{i-1}c_{ij}=1$, which implies the so-called Shu-Osher form.  
Thus, if the expression inside the square brackets in Eq.~\eqref{eq:imexStagesRewrite} --- which is in the form of a forward Euler update with time step $\hat{c}_{ij}\,\dt$ --- is positive for all $i=1,\ldots,s$, $j=0,\ldots,i-1$, positivity of the entire sum on the right-hand side of Eq.~\eqref{eq:imexStagesRewrite} follows from convexity arguments.  
Thus, if the explicit update with the transport operator is positive for a time step $\dt_{\mbox{\tiny Ex}}$, the IMEX scheme is positivity-preserving for a time step $\dt\le c_{\mbox{\tiny Sch}}\,\dt_{\mbox{\tiny Ex}}$, where
\begin{equation}
  c_{\mbox{\tiny Sch}}=\min_{\substack{i = 2,\ldots,s \\ 
       j = 1,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type A)}, \quad  c_{\mbox{\tiny Sch}}=\min_{\substack{i = 2,\ldots,s \\ 
              j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type ARS)}.
  \label{eq:imexCFL}
\end{equation}
It is desirable to make $c_{\mbox{\tiny Sch}}$ as large (close to $1$) as possible.  
(Positivity of $\vect{u}^{(i)}$ also requires the implicit solve to preserve positivity.)  
In \cite{hu_etal_2018}, Hu et al. provide examples of GSA, positivity-preserving IMEX schemes of type A (see scheme PA2 in \ref{app:butcherTables}) and type ARS.  
In \ref{app:butcherTables}, we provide another example of a GSA, positivity-preserving IMEX scheme of type A (scheme PA2+), with a significantly larger $c_{\mbox{\tiny Sch}}$ (a factor of about $1.7$ larger).
